\input{inputs/format_header.tex}
\setlength{\evensidemargin}{0pt}
\setlength{\oddsidemargin}{0pt}

\guidetitle{lasso workflow}{2015-11-10}

\begin{document}

\maketitle

\section{Example command}
All on one line. Before running with many samples, run with -samples=1 to check 
that everything works!!!
\begin{verbatim}
workflow -true_model=true.mod -base_model=base.mod -seed=7654
-external_dataset=large_data.dat -samples=1
\end{verbatim}
\section{Input}


\begin{description}
\item[-true\_model] The true model with all covariate effects, on log scale. 
\$DATA has the  small (20 individuals) template data set.
\item[-base\_model] The base model with no covariate effects. \$DATA has the 
small (20 individuals) template data set.
\item[-seed] A random seed
\item[-external\_dataset] A large dataset created by simulation from the true
model. This dataset must contain simulated observations (it cannot be the
template dataset)
\item[-samples] The number of small simulated datasets to run the lasso:s
with
\end{description}

\section{What the script does automatically}
\begin{enumerate}
\item The program will first simulate 'samples' small data sets using sse,
and then re-estimate the true and base models on each of the simulated data sets
\item Make 'samples' copies of the base model, set \$DATA to simulated 
data set $i$ from step 1, and update initial estimates to final estimates 
from re-estimation in step1
\item For each model from step 2, run
\begin{enumerate}
\item Adjusted adaptive lasso
\item Adaptive lasso (regular lasso is first iteration of this procedure)
\end{enumerate}
The options 
\begin{verbatim}
-retries=5 -picky -no-normalize -stop_t=4 -log_scale 
-relations=CL:COV1-2,COV2-2,COV3-2,COV4-2,COV5-2,COV6-2,
COV7-2,COV8-2,COV9-2,COV10-2
\end{verbatim}
will be set autmatically.
Coefficients from all samples and lasso variants will be collected in 
file final\_lasso\_coefficients.csv
\item For the three final models (aalasso, alasso, lasso) from step 3, create
evaluation model where \$DATA is set to external data set, \$EST is set to
MAXEVAL=0, and lasso coefficients are set to final coefficients. 
In total will get 3 times 'samples' evaluation models.
\item Make 'samples' copies of true model from step 1
set \$DATA to external data set, set MAXEVAL=0
update initial estimates to final estimates 
from re-estimation in step1. Will make 'samples' additional evaluation models.
\item Run all 4 times 'samples' evaluation models. Compare ofv and absolute 
prediction errors in file evaluation\_ofv.csv
\end{enumerate}

\end{document}
