\documentclass[a4wide,12pt]{article}
%\setlength{\marginparwidth}{0pt}%35
%\setlength{\marginparsep}{0pt}%?
%\setlength{\evensidemargin}{0pt}
%\setlength{\oddsidemargin}{0pt}
\usepackage{lmodern}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{verbatim}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage{alltt}
\usepackage{ifthen}
%\usepackage[backend=biber, maxbibnames=99, defernumbers]{biblatex}
%\addbibresource{PsN.bib}
% Reduce the size of the underscore
\usepackage{relsize}
\renewcommand{\_}{\textscale{.7}{\textunderscore}}

\newcommand{\revisiondate}[1]{
\date{#1}
}
\newcommand{\guidetitle}[2]{
\title{#1\\ }
\date{Revised #2}
}


\newcommand{\doctitle}[2]{
\title{#1}
\date{#2}
}

\newcommand{\optname}[1]{\item{{\bfseries\texttt-#1}\newline}}
\newcommand{\optdefault}[2]{\item{{\bfseries\texttt-#1}{\mbox{ = \it #2}}\newline}}

\guidetitle{m1find user guide}{2016-03-17}

\begin{document}

\maketitle
\newcommand{\guidetoolname}{m1find}


\section{Introduction}
\subsection{What is the m1 subfolder}
All PsN programs (except execute) that run NONMEM will create a m1 subfolder in
the main run directory. The m1 subfolder contains intermediate files,
such as bootstrapped data sets and lst-files from individual NONMEM runs, before
results are summarized. When a run is finished, the m1 folder is not cleaned
away by PsN, even though the contents are usually not needed.
%, because \emph{sometimes} the contents are still needed.

In most situations,
the m1 contents are just a nuisance, and the collected m1 subfolders
are a great burden to our file storage and backup system. The more files on
the system, the higher risk of file system or backup failure
and very long recovery times before we can work normally again.
About a third of all files on our system is old m1 folders, 
and if all
m1 folders were removed, we would not have a backup problem anymore.
In a few situations the m1 folders needs to be zipped and kept,
and those situations
are listed below.

\subsection{What is the NM\_run subfolder}
The NM\_run subfolder is a temporary folder where intermediate NONMEM output is created.
These folders are automatically removed by PsN except when you have changed the -clean setting or
when NONMEM or PsN crashes.
In the last case the contents of NM\_run may be used to diagnose what went wrong.
NM\_run subfolders should always be removed when the contents are not used for debugging.

\subsection{m1find}
The program m1find is a simple method to remove or zip m1 folders and/or to remove NM\_run folders. 
The names of all removed folders are logged in a file in the user's home directory,
and can be easily retrived from backup by the system administrator in case the
user realizes the folder is still needed.

\section{What to do with m1}
\subsection{When to keep m1 untouched}
The \emph{only} situation when m1 should be kept untouched is when it is part of a job that
is running on the cluster \emph{now}, i.e. something that shows
up in squeue. This is at most 20 m1 folders per user, unless you have many
models running in the lowprio queue.

{\bf All} other m1 subfolders should be zipped or, preferably, removed, and m1find can do it for you.

\subsection{When to zip m1}
\begin{itemize}
\item you are not done with binning in a vpc, then m1 contains files needed to redo the binning
\item you have run a lasso and you are not done analyzing the output models saved in m1  
\item you want to use -add\_retries in sse on previously simulated datasets \emph{and} simulated data sets take a
long time to regenerate. If data sets are quick to generate, then remove m1.
\item you have an interrupted bootstrap, where some but not all samples have lst-file in m1.
\emph{Note:} if you have a complete raw\_results file for
bootstrap you should remove m1, because you can recompute percentiles etc with different exclusion criteria (estimation\_terminated
and such options) using only the raw\_results file.
\end{itemize}
PsN will automatically unzip a zipped m1 folder if you are restarting a run in
an existing PsN run directory, 
so you do not have to unzip anything manually for example if you run sse with -add\_retries,
or redo the binning of a vpc.

\subsection{When to remove m1}
\begin{itemize}
\item Always except in cases listed above
\end{itemize}

\section{What to do with NM\_run}
\subsection{When to keep NM\_run}
When NONMEM (nmfe) is running in there now, or you are investigating the causes of a NONMEM/PsN crash.
\subsection{When to remove NM\_run}
All other cases. NM\_run folders will never be zipped by m1find.

\section{How to use m1find}

The command structure is\\
m1find ``what to do'' ``where'',\\
for example\\
m1find remove course\_work


\subsubsection*{where}
``where'' is the name of one or more directories, for example course\_work, Project\_1 or bootstrap\_dir2. Multiple directories
are listed with a space in between. A single dot (.) means the current working directory.
Unix wildcards can be used to easily input multiple directory names, for example sse* which will match
all directories in the current working directory that start with sse.
The program will find all m1/NM\_run/m1.zip \emph{anywhere inside} the given directories,
it can search through multiple directory levels. See examples further down.

\subsubsection*{what to do}
``what to do'' is either list, zip, remove or ask.
\begin{description}
\item[list] All m1 and NM\_run directories and m1.zip files anywhere inside the given folder(s) will be found
  and listed on screen. No zipping or removing will take place. 
\item[remove] All m1 and NM\_run directories and m1.zip files anywhere inside the given folder(s) will be found and listed on screen.
         The user will have the option to remove all.
         All removed files will be logged in the file removed\_by\_m1find in your home directory.
\item[zip] All m1 directories anywhere inside the given folder(s) will be found and listed on screen.
  The user will then have the option to zip all. Note that m1.zip files and NM\_run directories
  will be ignored with ``zip''.
\item[ask] All m1 and NM\_run directories and m1.zip files anywhere inside the given folder(s)
  will be found and presented one by one.
  The user has the option to remove, skip or, if it is a m1 directory, zip each item separately.
\end{description}

\subsubsection*{Do not disturb other cluster users}
To be on the safe side, always
start an interactive slurm session before running m1find,
to reduce the load on the login node.
\begin{verbatim}
srun --pty bash
\end{verbatim}

\noindent After you are done with m1find, end the session with command \verb|exit|.

\subsection{Examples}

\subsubsection*{List all my m1/NM\_run folders and m1.zip files}
Use cd to go to your home directory and then run
\begin{verbatim}
m1find list .
\end{verbatim}


\subsubsection*{Remove all m1 and NM\_run in project folder ProjectA}

\begin{verbatim}
m1find remove path/to/ProjectA
\end{verbatim}
Remember that names of all removed m1 and NM\_run will be stored in a file in your home directory.

\subsubsection*{Remove all NM\_run in project folder ProjectB (ignoring m1 and m1.zip)}

\begin{verbatim}
m1find remove path/to/ProjectB -no-m1
\end{verbatim}
Note the - at the beginning of -no-m1. If you skip it
then no-m1 will be interpreted as the name of a directory you want to go through. Remember that names of all removed NM\_run will be stored in a file in your home directory.

\subsubsection*{Keep a few m1 and NM\_run folders in a project directory}
I have 30 bootstrap folders in my project folder, where bootstrap\_dir 1-28 are done and
bootstrap\_dir29 and bootstrap\_dir30 are running now:
Remove all m1 and NM\_run in bootstrap\_dir with number starting with 1 (1, 10-19):
\begin{verbatim}
m1find remove bootstrap_dir1*
\end{verbatim}
Interactively go through all remaining m1/NM\_run, (want to remove all except
m1 and NM\_run in bootstrap\_dir29 and bootstrap\_dir30): 
\begin{verbatim}
m1find ask bootstrap_dir*
\end{verbatim}

\subsubsection*{Remove all m1 and NM\_run in ProjectA, except m1 with lasso results, and then zip all m1 with lasso results}

\begin{verbatim}
m1find remove path/to/ProjectA -skip_lasso_m1
m1find zip path/to/ProjectA
\end{verbatim}
Remember that names of all removed m1 and NM\_run will be stored in a file in your home directory.

\section{Additional options}
\begin{description}
\optname{no-interactive} Skip confirmation question before removing or zipping.
\optname{no-nm\_run} Skip all NM\_run directories. Do not set together with -no-m1, because then there is nothing to do for m1find. 
\optname{no-m1} Skip all m1 and m1.zip. Do not set together with -no-nm\_run or with ``zip'',
because then there is nothing to do for m1find. 
\optname{skip\_lasso\_m1} Skip m1 directories that contain output models from a lasso.
\optdefault{logdir}{directory} Default user's home directory. The directory to write
log file removed\_by\_m1find  
\optname{h} Print help text and exit.
\optname{help} Print help text and exit.
\end{description}

\end{document}
