\input{inputs/format_header.tex}
\guidetitle{BENCHMARK user guide}{2018-01-31}


\begin{document}

\maketitle
\newcommand{\guidetoolname}{benchmark}


\section{Introduction}
The benchmark tool is intended for systematic comparison of run times, ofv and parameter estimates
across NONMEM versions and/or alternative settings of NONMEM options.
For each model file given on the command line, PsN will create one copy for each possible combination
of the modifications listed on the command line (option -record\_options), and
then each model copy will be run first with the nonmem version given
with option -nm\_version (as with any PsN tool) and then also with
each of the nonmem versions listed with option -alt\_nonmem (if any).
Each model run will be repeated if option -replicates is set larger than 1.

There will be one raw\_results file produced for each input model, and each raw\_results file will contain 
'replicates' row(s)
for each combination of model modifications and NONMEM version. 
%The R plots functionality is used to
%read the raw\_results files and
%visualize the run times (as reported by NONMEM in the lst-files), the stability of ofv values and of estimates.

Note: The run times reported will be estimation step times, covariance step times and total times reported in the NONMEM lst-files.
PsN will not measure times, only parse the contents of the lst-file. It is up to the user to ensure that the run conditions 
(hard-ware, cpu load) are comparable across NONMEM runs (e.g. run on a homogeneous cluster where queueing system ensures 
no interference between NONMEM runs, or run with -threads=1 on standalone computer and no other significant runs at the same time).

Examples:
\begin{verbatim}
benchmark run1.mod run2.mod -nm_version=nm74alpha 
-record_options=estim:none,FAST -replicates=3

benchmark run3.mod -nm_version=nm73gfortran4 
-alt_nonmem=nm73gfortran5 -merge_rawreults -replicates=2
\end{verbatim}

See below for explanation of each option.

	\section{Input and options}

		\subsection{Required input}
		A model file is required on the command line. 
		Also, there must additional options resulting in at least two NONMEM runs that can be compared. This can be
		achieved by setting at least one alternative NONMEM version or at least one record modification listed,
		or giving at least two model files as input and setting option -merge\_rawresults. 
	
			\subsection{Optional input}
			\begin{optionlist}
			\optdefault{alt\_nonmem}{list}
			Default not set. 
			A comma-separated list of NONMEM versions, as named in psn.conf to run all, possibly modified, models with
			\emph{in addition to} the NONMEM version set with PsN common option -nm\_version.
			If different compilers for the same NONMEM version are to be compared, 
			then the same NONMEM version compiled with different compilers
			must be listed with different names in psn.conf, since PsN does not ''know'' about compilers.
			\nextopt
			\optname{copy\_data}
			Default set. Disable with -no-copy\_data. By default, PsN will copy the datafile into NM\_run1 and set a 
			local path in psn.mod, the actual modelfile run with NONMEM. If -no-copy\_data is set, PsN will not copy 
			the data to NM\_run1 and instead set a global path to the datafile in psn.mod. However, NONMEM will not 
			accept a path longer than 80 characters.
			\nextopt
			%\optdefault{dofv\_threshold}{number}
			%Default 1 (percent). If the relative difference between the ofv of two model runs in the same raw\_results file
			%is greater than this threshold, these models will be flagged in an extra column in the raw\_results file.
			%\nextopt
			%\optdefault{parameter\_threshold}{number}
			%Default 5 (percent). If the relative difference between the parameter estimates of two model runs 
			%in the same raw\_results file
			%is greater than this threshold, these models will be flagged in an extra column in the raw\_results file.
			%\nextopt
		    \nextopt
			\optname{merge\_rawresults}
			Default not set. The default is to create a separate raw\_results file for each input model,
			where results for all the variants resulting from -record\_options and -alt\_nonmem
			are collected. Creating separate raw\_results files for each input model is done because
			it cannot be assumed that the models and sets of population parameters are the same and comparable,
			so an investigation of the stability of e.g. parameter estimates across different input models
			is not meaningful and the headers in the raw\_results file can be completely wrong for some models. 
			However, the user can choose to override the default procedure
			with option -merge\_rawresults, resulting in all runs being reported in the same raw\_results file,
			and the R plots processing being performed for all models as a single group.
			\nextopt
			\optname{prepend\_model\_file\_name}
			Default not set. Set if you want to keep and compare table files.
			\nextopt
			\optdefault{record\_options}{list}
			Default not set.
			A structured list of the control stream records to modify, and the options to set. 
			The record names can be abbreviated and written in lower or upper case,
			as long as they are correct abbreviations of the complete record names. 
			For example: estimation, EST and estim are correct while ESIMATION
			is wrong. After the record name there must be colon (:), and then
			a comma-separated list of two or more alternative options.
			If only one alternative is set, this modification will be made to all models, 
			which only makes sense if there are other sets of alternative modifications.
			The option names must be in upper case, unless ``none'' is specified, which is 
			a marker for ``no modification of the record''. Example: \\
			-record\_options=estimation:none, FAST means that
			each model should be run both with option FAST added to \$ESTIMATION and no modification to 
			\$ESTIMATION.
			If more than one record-options combination is desired, the different combinations are 
			separated with double comma (,,). Example: \\
			-record\_options=est:none,FAST,,est:ATOL=10,ATOL=8,\\ATOL=12 
			which would result in six (2 times 3) variants of each model.
			\nextopt
			\optdefault{reference\_lst}{list}
			Add the results from one or more pre-existing .lst files to all raw\_results.
			Saves time if reference results already available, e.g. when comparing results
			from old NONMEM versions to a newly installed version.
			\nextopt
			\optdefault{replicates}{N}
			Default is 1. Can be set larger than 1 e.g. for run time comparisons.
			See note about user responsibility in ensuring run times are comparable.
			\nextopt
			\optdefault{theta\_inits}{list}
			Default not set.
			A structured list of the \$THETA initial estimates to modify. If used, the 
			\$THETAs in question will be modified regardless of whether they are FIXED in the control stream.
			In the list, the \$THETAs are identified by their order number, where numbering starts at 1,
			or by their PsN format label, as CL in this example:
			\begin{verbatim}
			$THETA (0,1.005) ; CL
			\end{verbatim}
			In the list, after the theta label/number there must be a colon and then a comma-separated list
			of numbers to set as initial estimates. If 'none' is set it means do not change the input model.
			If an additional theta is to be modified, then a double comma separates 
			the two sub-lists. If no \$THETA pre-exists with the label given in the list,
			a \$THETA will be added with that initial estimate and label. 
			If \$THETA number is used but no \$THETA exists with that number then benchmark will halt with an
			error message.
			If the same theta is referred to with both label and theta number, this will not be detected
			by benchmark and will result in incorrect results.
			Example, using both label and theta number
			\begin{verbatim}
			-theta_inits=CL:none,1.5,1.0,,5:8.5,12
			\end{verbatim}
			\nextopt
			\optdefault{threads}{integer}
			Default is 5 (if the default psn.conf is used). Use the threads option to enable parallel execution of multiple models. This option decides how many models PsN will run at the same time, and it is completely independent of whether the individual models are run with serial NONMEM or parallel NONMEM. If you want to run a single model in parallel you must use options -parafile and -nodes. On a desktop computer it is recommended to not set -threads higher the number of CPUs in the system plus one. 
			
			You can specify more threads, but it will probably not increase the performance. If you are running on a computer cluster, you should consult your system administrator to find out how many threads you can specify.
			
			This option must be set carefully in benchmark if the run time of models is to be investigated. Comparison of run times will be meaningless if the individual NONMEM runs are not performed under the same conditions. 
			\nextopt
						\end{optionlist}

\subsection{PsN common options}
For a complete list see common\_options.pdf or type psn\_options -h on the command line.

\section{Output}
The output from benchmark is
%, in addition to the plots,
one raw\_results file for each input model, or all input models together if -merge\_rawresults was set.
In each file the first model run will be used as reference when computing differences in output.
The raw\_results file(s) will have extra columns for the input model name, each
modification set via -record\_options, the NONMEM version if -alt\_nonmem was used, and the replicate
number if -replicates was set.
%\begin{itemize}
%\item delta-ofv (dofv) relative to the reference run
%\item delta-runtime 
%\item delta-estimates 
%\end{itemize}


\end{document}

\subsection{Auto-generated R plots from PsN}
\newcommand{\rplotsconditions}{The default benchmark template 
requires no special R libraries.
If no pdf is generated,
see the .Rout file in the main run directory for error messages.}
\input{inputs/rplots_section_body.tex}

\subsubsection*{Basic R plots}
Basic diagnostic R plots for each input model, or all input models together if -merge\_rawresults was set, 
will be generated if option -rplots is set >0.
The basic R plots show differences in total run time, estimation time, covariance time and ofv.
If the ofv-difference between any two runs is greater than a threshold (see option -dofv\_threshold), this is flagged. 

\subsubsection*{Extended R plots}
Extended diagnostic R plots for each input model, or all input models together if -merge\_rawresults was set, 
will be generated if option -rplots is set >1.
The extended R plots show differences in parameter estimates and standard errors.
If the estimate difference is greater than a threshold (see option -parameter\_threshold) then this is flagged.
