\input{inputs/format_header.tex}
\guidetitle{EBE\_NPDE user guide}{2015-11-16}

\begin{document}

\maketitle
\newcommand{\guidetoolname}{ebe\_npde}


\section{Overview}

The ebe\_npde program is a tool for simulation-evaluation based diagnostics.
%based on objective functiuon value (OFV), individual function value (iOFV)
%and empirical Bayes estimates (EBEs). 
The principle is 
\begin{enumerate}
\item Simulate N data sets from final model
\item Apply final model to each data set (evaluate given the model parameters used for simulation or evaluate with model parameters that are re-estimated for each simulated data set)
\item Extract output for diagnostics
\begin{description}
\item[OFV] population
\item[iOFV] individual
\item[EBE] parameter
\item[IWRES] residual
\end{description}
\item Provide reference distributions for ”observed” OFV, iOFV, EBE and IWRES
\end{enumerate}
The program computes normalized prediction distribution errors for EBE:s and IWRES 
using the procedure described in \cite{Comets}, with example use in \cite{Keizer} and \cite{Largajolli}.
It also performs OFV diagnostics. Diagnostics plots are created via the -rplots option, see section 
Auto-generated R-plots from PsN.

Example call
\begin{verbatim}
ebe_npde run1.mod -rplots=1
\end{verbatim}

\section{Input and options}

\subsection{Required input}

A model file is required on the command-line.

\subsection{Optional input}

\begin{optionlist}
\optdefault{lst\_file}{filename}
Default not set. By default PsN will, before copying the input model to the simulation models, look for an output file with final estimates for the input model and if found update the initial estimates in the simulation models. If option -lsf\_file is set Psn will instead use the final estimates in the given file. If option is not set and no output file is found PsN will not update the estimates but instead rerun the input model, if option -estimate\_input is set. 
\nextopt
\optname{estimate\_input}
Default set. By default, PsN will rerun the input model to get parameter estimates unless an output file is found or option -lst\_file is set. But if option -estimate\_input is unset with -no-estimate\_input and no lst-file is found then the parameter estimates set in the input model will be used for simulations. 
\nextopt
\optdefault{samples}{N}
Default 300. The number of simulations and reestimations/evaluations to run. 
\nextopt
\optdefault{n\_simulation\_models}{N}
Default equal to the number of threads.
By setting this option to a number N greater than 1, the 'samples' simulations and evaluations
will be split equally between N model files, which can be run in parallel.
\nextopt
\optname{reminimize}
Default not set. By default, simulated datasets will be run with MAXEVAL=0 (or equivalent for non-classical estimation methods). If option -reminimize is set then \$EST will be the same as in the input model. 
\nextopt
\optdefault{gls\_data\_file}{filename}
Default gls\_data.dta. A file with input data for the gls program is always generated. This option changes the name of that file. After this file is created, the gls program can be run with option \mbox{-gls\_model} and -ind\_shrinkage with minimum runtime, see gls\_userguide.pdf 
\nextopt
%\optdefault{iov}{list}
%TODO: list of iov etas, to be treated from all other etas assumed to be iiv. 
%\nextopt
\end{optionlist}

\subsection{Some important common PsN options}
For a complete list see common\_options.pdf, 
or psn\_options -h on the commandline.
\input{inputs/basic_options.tex}

\subsection{Auto-generated R-plots from PsN}
\newcommand{\rplotsconditions}{The default ebe\_npde template 
requires that library PerformanceAnalytics is installed.
If the conditions are not fulfilled then no pdf will be generated,
see the .Rout file in the main run directory for error messages.}
\input{inputs/rplots_section_body.tex}

%\subsubsection*{Basic plots}

\section{Output}

\subsubsection*{Main output}
\begin{description}
\item[ebe\_npde.csv] NPDE of EBE. One row per subject. Columns: ID,
standardized observed EBE (a scalar) where 
$ebe_{standardized}= \\
\left(\overline{ebe}_{obs} - mean (\overline{ebe}_{sim})\right)\left((cov(\overline{ebe}_{sim})\right)^{-1}
\left(\overline{ebe}_{obs} - mean (\overline{ebe}_{sim})\right)^T$,\\ 
NPDE:s for each EBE
\item[summary\_iwres.csv] NPDE of IWRES. One row per data record. Columns: ID, MDV, observed IWRES, NPDE for IWRES.
\item[summary\_iofv.csv] One row per subject. Columns: ID, observed iOFV,
mean of iOFV for evaluation of simulated data,
standard deviation of iOFV for evaluation of simulated data,
standardized observed iOFV where
$iOFV_{standardized}=\frac{(iOFV_{observed} - mean (iOFV_{simulated}))^2}{variance(iOFV_{simulated})}$, NPDE of iOFV
\item[raw\_all\_iofv.csv] One row per subject. Columns: ID, observed iOFV, sample.1 iOFV, sample.2 iOFV, etc. 
\item[raw\_results\_<model>.csv] e.g. raw\_results\_run1.csv with ofv for evaluation of simulated data
\end{description}

\subsubsection*{Additional output}
\begin{description}
\item[decorrelated\_original\_ebe.csv] One row per individual. Columns: ID, decorrelated observed EBE:s
\item[decorrelated\_original\_iwres.csv] One row per data record. Columns: ID, MDV, decorrelated observed IWRES
\item[raw\_original\_ebe.csv]  One row per individual. Columns: ID, observed EBE:s
%\item[raw\_original\_iwres.csv] One row per data record. Columns: ID, MDV, observed IWRES %skip, in summary
\item[ebe\_pde.csv] One row per individual. Columns: ID, PDE of EBE:s
\item[ebe\_npd.csv] One row per individual. Columns: ID, NPD of EBE:s (normalized PD, not decorrelated)
\item[iwres\_npd.csv]  One row per data record. Columns: ID, MDV, NPD of IWRES (normalized pd, not decorrelated)
\end{description}

\section{Algorithm overview}

\begin{enumerate}
\item Remove MSFO option from \$EST.
\item Read model file to find set of IIV ETAs. An ETA is assumed to be IIV if the corresponding \$OMEGA is not SAME
\emph{and} the next \$OMEGA is not SAME either. Also find set of IOV ETAs, grouped by occasions.
An ETA is assumed to belong to IOV occasion 1 if the corresponding \$OMEGA is not SAME, but the next \$OMEGA is SAME. 
An ETA is assumed to belong to IOV occasion 1+$i$ if the corresponding \$OMEGA the $i$th consecutive SAME \$OMEGA. 
\item If \$SIM not present, create simple \$SIM as in sse. Create 'n\_simulation\_models' copies of modified
input model with order number of copy indicated in filename. In each copy set unique seed in \$SIM and set
NSUB so that sum of simulations is equal to 'samples', remove old \$TABLE if present, 
if option reminimize is not set then set MAXEVAL=0, and
add \$TABLE IWRES ID NOPRINT ONEHEADER NOAPPEND FILE=iwres-$\langle$ordernum$\rangle$.dta. 
\item In 'original' input model: Remove \$SIM if present. Add \$TABLE $\langle$all undropped items in \$INPUT$\rangle$ IPRED PRED IWRES NOPRINT ONEHEADER NOAPPEND FILE=orig\_pred.dta. Add \$TABLE IWRES ID MDV NOPRINT ONEHEADER NOAPPEND FILE=original\_iwres.dta. 
\item run modified original input model and 'n\_simulation\_models' sim models.
\item Read all iwres\_$\langle$order number$\rangle$.dta files,
storing IWRES values per data point. Compute, per data point, ISHR\_ij=1-sd(IWRES\_ij). Open orig\_pred.dta, append ISHR column with
computed values, and print to gls\_input.dta. Print also shrinkage column to new file ind\_iwres\_shrinkage.dta. File gls\_input.dta
can be used as input when running gls program, see gls\_userguide.pdf. 
\item Compute npde for IWRES according to \cite{Comets}.
\item Read ETA values from 
phi-files for original and simulated data, and compute npde for EBE:s according to \cite{Comets}.
Perform computations for IIV, and also for IOV for different occasions, separately. 
\end{enumerate}

\begin{thebibliography}{99}
\bibitem{Comets} E. Comets, K. Brendel and F. Mentré,
{\em Computing normalised prediction distribution errors to evaluate nonlinear mixed-effect models: The npde add-on package for R},
CMPB 90 (2008) 154–166.
\bibitem{Keizer} R. Keizer, K. Harling and M. O. Karlsson,
{\em Extended NPDE diagnostics for the between-subject variability and residual error model},
PAGE 21 (2012) Abstr 2538 \mbox{www.page-meeting.org/?abstract=2538}.
\bibitem{Largajolli} A. Largajolli, S. Jönsson and M. O. Karlsson,
{\em The OFVPPC: A simulation objective function based diagnostic}
PAGE 23 (2014) Abstr 3208 \mbox{www.page-meeting.org/?abstract=3208}.
\end{thebibliography}     


\end{document}

