\input{inputs/format_header.tex}
\guidetitle{EXECUTE user guide}{2017-10-12}


\begin{document}

\maketitle
\newcommand{\guidetoolname}{execute}


\section{Introduction}
The execute program is a PsN tool that allows you to run multiple modelfiles either sequentially or in parallel. 
It is an nmfe replacement with advanced extra functionality.

The execute program creates subdirectories where it puts NONMEMs input and output files, to make sure that parallel 
NONMEM runs do not interfere with each other. The top directory is by default named 'modelfit\_dirX' where 'X' is a 
number that starts at 1 and is increased by one each time you run the execute program. When the NONMEM runs are 
finished, the output and table files will be copied to the directory where execute started in which means that you can 
normaly ignore the 'modelfit\_dirX' directory. If you need to access any special files you can find them inside the  
'modelfit\_dirX'. Inside the 'modelfit\_dirX' you find a few subdirectories named 'NM\_runY'. For each model file you 
specified on the command line there will be one 'NM\_runY' directory in which the actual NONMEM execution takes place. 
The order of the 'NM\_runY' directories corresponds to the order of the modelfiles given on the command line. The first run 
will take place inside 'NM\_run1', the second in 'NM\_run2' and so on. The execute script is a PsN tool that allows you to run 
multiple modelfiles either sequentially or in parallel. It is an nmfe replacement with advanced extra functionality.\\
Examples:\\
Run one model file and accept all default values.

\begin{verbatim}
execute pheno.mod
\end{verbatim}

Run two model files in parallell using 5 possible retries:

\begin{verbatim}
execute -threads=2 -retries=5 run1.mod 
\end{verbatim}

\section{Output}
When the NONMEM runs are finished, output and table files will be copied to the directory in which execute started which means 
that you normaly can ignore the run directory. If you need to access any special files, you can find them inside the run directory. 
The run directory also contains a raw\_results csv-file, a comma-separated file with summary information about minimization status, 
run time, ofv, parameter estimates etc. In the run directory you find a few subdirectories named 'NM\_runY'. For each model file 
specified on the command line there will be one 'NM\_runY' directory in which the actual NONMEM execution takes place. 
The order of the 'NM\_runY' directories corresponds to the order of the modelfiles given on the command line. The first run will take 
place inside 'NM\_run1', the second in 'NM\_run2' etc.

\section{Required input}
A model file is required as argument.
	
\section{Optional input}
All options listed in the document common\_options.pdf, apply to execute. Those options govern how NONMEM runs are managed, 
and hence apply to all PsN scripts that run NONMEM. A large selection of those options are included in this document,
together with the options that only apply to execute. 
		
\renewcommand{\guidetoolname}{modelfit}
		
\subsection{Basic options}
The following options are the ones most commonly used. 
\input{inputs/basic_options.tex}

\renewcommand{\guidetoolname}{execute}
			
\begin{optionlist}
\optname{boxcox}
Set to boxcox transform all ETAs before running. Cannot be used on models with multiple \$PROBLEMs.
\nextopt
\optname{copy\_data}
Default set.  By default, PsN will copy the datafile into NM\_run1 and set a local path in psn.mod, the actual modelfile run with NONMEM. Disable with -no-copy\_data. If -no-copy\_data is set, PsN will not copy the data to NM\_run1 and instead set a global path to the datafile in psn.mod.NONMEM will not accept a path longer than 80 characters.
\nextopt
\optname{etas}
If available, use the phi-file of the previous run as initial estimates for the etas. The phi-file will be used in \$ETAS.
\nextopt
\optname{model\_dir\_name}
Default not set. This option changes the basename of the run directory from modelfit\_dir to $\langle$modelfile$\rangle$.dir, where $\langle$modelfile$\rangle$ is the name of the (first) input model file, without the extension. This option is specific to execute.
The directories will be numbered starting from 1, increasing the number each time execute is run with a model file with the 
same name. If the option directory is used it will override -model\_dir\_name.
\nextopt
\optname{timestamp}
Default not set. This option changes the name of the run directory to $\langle$modelfile$\rangle$-PsN-$\langle$date$\rangle$
where $\langle$modelfile$\rangle$ is the name of the first model file in the list given as arguments, without the extension,
and $\langle$date$\rangle$ is the time and date the run was started. 
Example: directory name run1-PsN-2014-06-12-152502 for a run that was started at 15:25:02 June 12th in year 2014.
This option is specific to execute.
If the option directory is used it will override -timestamp.
\nextopt
\end{optionlist}

\subsection{Auto-generated R-plots from PsN}
\newcommand{\rplotsconditions}{The default execute template 
requires the xpose4 R library. It also relies on Xpose-type tables being 
created, such as sdtab, patab and cotab, with the correct run number. 
See the Xpose documentation for requirements on such table files,
e.g. pages 106-108 in\\
\texttt{http://xpose.sourceforge.net/bestiarium\_v1.0.pdf}.\\
If option -subset\_variable\_rplots is used, 
the user must ensure that the subset variable
is printed to one of the xpose tables, for example sdtab, and then
there will be separate plots created for
subsets of the data, via xpose options 'subset' and 'by'. 
Xpose will treat the subset variable as continuous or categorical
based the number of unique values. The template contains 
a comment line with the xpose command for changing the default classification. 

If the conditions for creating the plots are not fulfilled then no pdf will be generated,
see the .Rout file in the main run directory for error messages.}
\input{inputs/rplots_section_body.tex}

\subsubsection*{Basic plots}
A pdf-file with 
basic execute rplots will be generated if option -rplots is set >0,
and the general rplots conditions fulfilled, see above.
The basic plots are created with the xpose4 functions
basic.gof, ranpar.hist, ranpar.qq and a few others. Please
refer to the xpose4 documentation for details regarding these
plots.

\subsubsection*{Extended plots}
Extended execute rplots will be generated if option -rplots is set >1.
These are individiual plots of ten randomly chosen IDs.

\subsection{Retries}
\input{inputs/retries_section_body.tex}

\subsection{Output handling and processing}
\input{inputs/output_processing.tex}
\begin{optionlist}
\optname{prepend\_options\_to\_lst}
Default not used. This option makes PsN prepend the final lst-file (which is copied back to the directory from which execute was called) with 
the file version\_and\_option\_info.txt which contains run information, including     all actual values of optional PsN options. PsN can still parse 
the lst-file with the options prepended, so the file can still be used it as input to e.g. sumo, vpc or update\_inits. Disabled with 
-no-prepend\_options\_to\_lst if set in psn.conf.
\nextopt
\end{optionlist}

\subsection{Running the nmfe script}
PsN always runs NONMEM via the nmfe script (or NMQual). 
The following options govern input for
nmfe, allowing the user to set options, such as \mbox{-parafile}, on the PsN commandline that
are passed on unchanged to the nmfe script.
Please refer to the NONMEM documentation for the meaning and appropriate use
of the nmfe script options.
\input{inputs/nmfe_interface.tex}

\subsection{Run progress messages}
\input{inputs/run_info_and_monitoring.tex}
\begin{optionlist}
\optname{tail\_output}
Default not used. This option only works for execute under Windows. Option -tail\_output specifies that execute should invoke a program (tail) 
that displays the output file, including the gradients, during minmization. The tail program is started automatically but it is up to the user to 
terminate the program. For the tail\_output option to work, a third party tail program must be installed. Tail programs that are known to work 
are WinTail and Tail for Win32.  The latter is recommended and can be downloaded at http://tailforwin32.sourceforge.net. It is also necessary 
to have correct settings of the variables wintail\_exe, which is the path to the tail program, and wintail\_command, which is the command for 
the tail program. The defaults, which work for the Tail for Win32 package, are seen below. These two variables can be set to other values in 
psn.conf.
\nextopt
\optdefault{wintail\_exe}{string}
Default \verb|C:\Program Files\Tail-4.2.12\Tail.exe|. Only for Windows. See tail\_output for description.
\nextopt
\optdefault{wintail\_command}{string}
Default "tail OUTPUT”. Only for Windows. See -tail\_output for description.
\nextopt
\end{optionlist}

\subsection{Control stream manipulation}
\input{inputs/control_stream_manipulation.tex}
\subsubsection{Transform Both Sides}
\input{inputs/tbs_options.tex}

\subsection{Error handling}
\input{inputs/error_checking_and_management.tex}


\subsection{Advanced execute options}
\begin{optionlist}
\optname{predict\_data}{string}
Only allowed in combination with -predict\_model.
Create a copy of the model with new file named predict\_model, turn off estimation 
(MAXEVAL=0 or corresponding), change data file to -predict\_data and then run.
\nextopt
\optname{predict\_model}{string}
Only allowed in combination with -predict\_data.
Create a copy of the model with new file named predict_model, turn off estimation (MAXEVAL=0 or corresponding), change data file to 
-predict\_data and then run.
\nexopt

\end{optionlist}

\references

\end{document}
