\input{inputs/format_header.tex}
\guidetitle{RESMOD user guide}{2021-05-25}
\usepackage{hyperref}
\begin{document}

\maketitle
\newcommand{\guidetoolname}{resmod}
\tableofcontents
\newpage

\section{Introduction}
The resmod (residual modelling) tool performs modelling of the conditional weighted residual output\cite{Ibrahim} of a model run. Different types of models are run and changes in OFV will be recorded together with some parameters of interest. The resmod tool is not compatible with NONMEM 7.2 or older versions.\\
Example:
\begin{verbatim}
resmod run1.mod
\end{verbatim}

\section{Input and options}

\subsection{Required input}
A model file is required on the command line.

\subsection{Optional input}

\begin{optionlist}
\optdefault{dv}{dependent variable}
Default is CWRES. Name of the dependent variable. 
\nextopt
\optdefault{dvid}{DVID column name}
Default is DVID. Name of the dvid column. 
\nextopt
\optdefault{groups}{N}
Default is 4. Set the number of groups to use for the time varying models. Quantiles using this number will be calculated.
\nextopt
\optname{idv}
Default is TIME. Name of the independent variable. If set to PRED PsN will rename it appropriately so that it can be used in \$INPUT.
\nextopt
\optname{iterative}
Set if an iterative process is wanted. The iteration is to pick the model with best dOFV and rerun all models with the CWRES output from that model etc until all models have been picked.
\nextopt
\optname{max\_iterations}
Default is no limit. Limit the maximum number of steps in the iterative procedure. 
\nextopt
\optname{obs\_column}
Default is the same as -dv. Use this column to ignore non-observations. All zero entries in this column will be interpreted as a non-observation.
\nextopt
\optname{occ}
Default is OCC. Name of the occasion column. 
\nextopt
\end{optionlist}

\section{Iteration}
If the option -iterative has been selected the model with the lowest dOFV will be selected. The output from this model
will be used to rerun all the models once more etc. The iteration will stop when either the -max\_iterations has been
reached or when all models have been selected. The iteration will be done for each DVID separately.

\section{Output}
The delta OFVs of all models are summarized in the resmod\_results.csv file together with the interesting parameter estimates.

\section{Residual models}

\subsection{Base model 1}
This model is used as a base model for the eta-on-epsilon, power, time-varying and the AR models.

\begin{verbatim}
Y = THETA(1) + ETA(1) + ERR(1)
\end{verbatim}

\subsection{IIV on RUV}
This model is compared to Base model 1 and is named \verb|IIV_on_RUV| in resmod\_results.csv. The \%CV for the omega for ETA(2) is presented.

\begin{verbatim}
Y = THETA(1) + ETA(1) + ERR(1) * EXP(ETA(2))
\end{verbatim}

\subsection{Power}
This model is compared to Base model 1 and is named \verb|power| in resmod\_results.csv. The power (THETA(2)) is presented.
If IPRED is not available in the original model the CIPREDI will be used instead.

\begin{verbatim}
Y = THETA(1) + ETA(1) + ERR(1)*(IPRED)**THETA(2)
\end{verbatim}

\subsection{Time varying RUV}
This model is compared to Base model 1 and is named \verb|time_varying_RUV| in resmod\_results.csv. The <idv> below will be replaced with the name of the independent variable. The <q1>, <q2> and <q3> below will be replaced with the quartiles of the independent variable for all observations combined. I.e. from the data set remove all records for which CWRES = 0 then calculate the quartiles of the idv column. The standard deviations of the epsilons and the time cutoff values are presented in the resmod\_results.csv.
\begin{verbatim}
Y = THETA(1) + ETA(1) + ERR(4)
IF (<idv>.LT.<q1>) THEN
    Y = THETA(1) + ETA(1) + ERR(1)
            'END IF
IF (<idv>.GE.<q1> .AND. <idv>.LT.<q2>) THEN
    Y = THETA(1) + ETA(1) + ERR(2)
END IF
IF (<idv>.GE.<q2> .AND. <idv>.LT.<q3>) THEN
    Y = THETA(1) + ETA(1) + ERR(3)
END IF
\end{verbatim}

\subsection{Autocorrelation}
This model is compared to Base model 1 and is named \verb|autocorrelation| in resmod\_results.csv. The autocorrelation (THETA(2)) will be presented. TIME will always be used as idv even if the option -idv is specified.
\begin{verbatim}
"FIRST
" USE SIZES, ONLY: NO
" USE NMPRD_REAL, ONLY: C=>CORRL2
" REAL (KIND=DPSIZE) :: T(NO)
" INTEGER (KIND=ISIZE) :: I,J,L
"MAIN
"C If new ind, initialize loop
" IF (NEWIND.NE.2) THEN
"  I=0
"  L=1
"  OID=ID
" END IF
"C Only if first in L2 set and if observation
"C  IF (MDV.EQ.0) THEN
"  I=I+1
"  T(I)=TIME
"  IF (OID.EQ.ID) L=I
"
"  DO J=1,I
"      C(J,1)=EXP((-0.6931/THETA(2))*(TIME-T(J)))
"  ENDDO
Y = THETA(1) + ETA(1) + EPS(1)
\end{verbatim}
\newpage

\subsection{T-distribution base model}
This model is used as base model for the t-distribution model.

\begin{verbatim}
    IPRED = THETA(1) + ETA(1)
	W     = THETA(2)
	IWRES=(DV-IPRED)/W
	LIM = 10E-14
	IF(IWRES.EQ.0) IWRES = LIM
	LL=-0.5*LOG(2*3.14159265)-LOG(W)-0.5*(IWRES**2)
	L=EXP(LL)
	Y=-2*LOG(L)
\end{verbatim}

Using the Laplace estimation method.

\subsection{T-distribution 2LL}
Named \verb|tdist_2ll_dfest| in resmod\_results.csv. The degree of freedom is estimated and presented in resmod\_results.csv.

\begin{verbatim}
IPRED = THETA(1) + ETA(1)
W = THETA(2)
DF = THETA(3) ; degrees of freedom of Student distribution
SIG1 = W ; scaling factor for standard deviation of RUV
IWRES = (DV - IPRED) / SIG1
; Nemesapproximation of gamma funtion(2007) for
; first factor of t-distrib(gamma((DF+1)/2))
PHI = (DF + 1) / 2
INN = PHI + 1 / (12 * PHI - 1 / (10 * PHI))
GAMMA = SQRT(2 * 3.14159265 / PHI) * (INN / EXP(1)) ** PHI
; Nemesapproximation of gamma funtion(2007) for
; second factor of t-distrib(gamma(DF/2))
PHI2 = DF / 2
INN2 = PHI2 + 1 / (12 * PHI2 - 1 / (10 * PHI2))
GAMMA2 = SQRT(2*3.14159265/PHI2)*(INN2/EXP(1))**PHI2
; coefficient of PDF of t-distribution
COEFF=GAMMA/(GAMMA2*SQRT(DF*3.14159265))/SIG1
BASE=1+IWRES*IWRES/DF ; base of PDF of t-distribution
POW=-(DF+1)/2 ; power of PDF of t-distribution
L=COEFF*BASE**POW ; PDF oft-distribution
Y=-2*LOG(L)
\end{verbatim}

\subsection{DTBS base model}
This model is used as base model for the DTBS model. If IPRED is not available in the original model the CIPREDI will be used instead.
\begin{verbatim}
    IPRT   = THETA(1)*EXP(ETA(1))
	WA     = THETA(2)
	LAMBDA = THETA(3)
	ZETA   = THETA(4)
	IF(IPRT.LT.0) IPRT=10E-14
	W = WA*IPRED**ZETA
	IPRTR = IPRT
	IF (LAMBDA .NE. 0 .AND. IPRT .NE.0) THEN
		IPRTR = (IPRT**LAMBDA-1)/LAMBDA
	ENDIF
	IF (LAMBDA .EQ. 0 .AND. IPRT .NE.0) THEN
		IPRTR = LOG(IPRT)
	ENDIF
	IF (LAMBDA .NE. 0 .AND. IPRT .EQ.0) THEN
		IPRTR = -1/LAMBDA
	ENDIF
	IF (LAMBDA .EQ. 0 .AND. IPRT .EQ.0) THEN
		IPRTR = -1000000000
	ENDIF
	IPRT = IPRTR
	Y = IPRT + ERR(1)*W
	IF(ICALL.EQ.4) Y=EXP(DV)
\end{verbatim}

Lambda and zeta will be fixed to 0.

\subsection{DTBS}
Named \verb|dtbs| in resmod\_results.csv. The lambda and zeta parameters (THETA(3) and THETA(4)) will be presented. If IPRED is not available in the original model the CIPREDI will be used instead. This model is same as the base model except that all parameters will be estimated.

\subsection{L2}
Name \verb|L2| in resmod\_results.csv. The L2 model will be tried only if the original model has multiple DVs. The L2 data column will be taken to be a copy of the TIME column. Resulting delta ofv will be the difference between the sum of the OFVs of the base models for all DVIDs and the OFV of the L2 model.
\begin{verbatim}
IF(DVID.EQ.1) Y = THETA(1) + ETA(1) + ERR(1)
IF(DVID.EQ.2) Y = THETA(2) + ETA(2) + ERR(2)
\end{verbatim}

\references

\end{document}
